Viscous Erosion in Porous Media

### 2018-01-10 docs/methods/erosion_methods.tex
Put Gaussian filter in paper

The constant of proportionality sets the time scale, and we set its
value to be 1
\begin{align}
  \dot{\xx}(t) = \abs{\tau} \nn.
  \label{eqn:bodyEvolution}
\end{align}

@@

% Quasi-steady comment
%We use a quasi-static approximation so that the velocity field instantaneously changes with the changing geometry.  Therefore, we can alternate between the two main aspects of our simulation
%\begin{enumerate}
%\item Given a geometry, compute the shear stress of along the boundary of each of the bodies.
%\item Given the shear stress, update the geometry according to the erosion model.
%\end{enumerate}

@@

For the sake of numerical stability, we will regularize the boundary evolution in two ways. First, we will smooth the absolute

In order to regularize boundary evolution, we modify the normal velocity
in \eqref{eqn:bodyEvolution} as follows
\begin{equation}
\label{eqn:Vn}
\Vn = \atau + \eps \mean{\atau} \left( \frac{L }{2\pi} \kappa - 1 \right)
\end{equation}
Here, $\kappa$ is the curvature, $\eps$ is a small parameter, and $L$ is the total arc length of the body, and $\mean{\cdot}$ indicates a surface mean. The second term on the right acts to smooth bodies with high-frequency oscillations by introducing a {\em curvature}-dependent component in the interface velocity. Since a closed body has a mean curvature of $\mean{\kappa} = 2\pi / L$, the term in parentheses has mean zero. Thus, this term is area-preserving, so that the only material loss is due to the shear stress itself. In addition, we have scaled this term on the $\mean{\atau}$, the average absolute stress, so that it is always proportional to the true source of erosion $\atau$, even as bodies vanish (reword).

@@

With the shear stress computed, we are now ready to evolve the interface of each body. As described in Section \ref{sec:thetaL}, we use the {\thL} framework, with $\theta$ being the tangent angle and $L$ the total arc length of a given boundary.

First, we introduce the variables
\begin{align}
\label{eq:M}
& \Mterm = - 2 \pi \mean{\thalpha \Vn} \\
\label{eq:N}
& \NLterm = \frac{2 \pi}{L} \left( \pderiv{\atau}{\alpha} + \thalpha \Vs \right) \\
\label{eq:elfun}
& \elfun = \frac{2 \pi  \mean{\atau}}{L}
\end{align} 
so that evolution equations \eqref{eqn:Levo1} and \eqref{eqn:thetaevo1} can be compactly expressed as
\begin{align}
\label{eqn:Levo2}
& \tderiv{L}{t} = \Mterm \\
\label{eqn:thetaevo2}
& \pderiv{\theta}{t} = \eps \elfun \ppd{\theta}{\alpha} + \NLterm
\end{align}
Here $\NLterm$ represents the nonlinear terms (and nonlocal etc) in the PDE for $\theta$.
We will evolve $\theta$ in spectral space, and so introduce the transforms
\begin{align}
& \theta(\alpha,t) = \alpha + \FourierSum \thhat_k e^{ i k \alpha} \\
& \NLterm(\alpha,t)  = \FourierSum \hat{\NLterm}_k e^{ i k \alpha}
\end{align}
Recall that $\alpha = 2 \pi s / L$ is a normalized arc length. In spectral space, \eqref{eqn:thetaevo2} produces the system of ODEs
\begin{equation}
\tderiv{\thhat_k}{t} +  \eps k^2  \elfun \thhat_k = \hat{\NLterm}_k
\end{equation}
Each ODE is first-order and linear (though variable-coefficient due to $\elfun$) and thus can be solved by using the integrating factor
\begin{equation}
\label{eq:mu}
\mu_k = \exp \left( \eps k^2 \int \elfun \, dt \right)
\end{equation}
We are therefore led to the coupled system of ODEs
\begin{align}
\label{eqn:Levo2}
& \tderiv{L}{t} = \Mterm \\
\label{eqn:thetaevo2}
& \tderiv{}{t}\left( \mu_k \thhat_k \right) = \mu_k \hat{\NLterm}_k
\end{align}
We solve system \eqref{eqn:Levo2}--\eqref{eqn:thetaevo2} with an explicit, second-order Runge-Kutta method (RK2). In particular, we choose the {\em midpoint} RK2 method so as to maximize the smoothing effect of the curvature term.

@@

Next, at the second stage of RK2, we discretize as
\begin{align}
& \frac{L^{(n+1)} - L^{(n)}}{\Dt} = \Mterm^{(n+1/2)} \\
& \frac{ \mu_k^{(n+1)} \thhat_k^{(n+1)} - \thhat_k^{(n)}}{\Dt} = \mu_k^{(n+1/2)} \hat{\NLterm}_k^{(n+1/2)}
\end{align}
We again compute the integrating factor, $\mu_k^{(n+1)}$ via the ( insert integration rule), giving
\begin{align}
\thhat_k^{(n+1)} = 
& \thhat_k^{(n)}
\exp \left( - \epsilon k^2 \Dt \elfun^{(n+1/2)} \right) +  \\
& \Dt \, \hat{\NLterm}_k^{(n+1/2)}
\exp \left( - \frac{1}{4} \eps k^2 \Dt \left( 3 \elfun^{(n+1/2)} - \elfun^{(n)} \right) \right)
\end{align}
\todo[inline]{Gaussian Filter}

>>>>

The constant of proportionality sets the time scale, and we set its
value to be 1
\begin{align}
  \dot{\xx}(t) = \abs{\tau} \nn.
  \label{eqn:bodyEvolution}
\end{align}
For the sake of numerical stability, we modify this law in Section \ref{sec:thetaL}.

@@

%We use a quasi-static approximation so that the velocity field instantaneously changes with the changing geometry.  Therefore, we can alternate between the two main aspects of our simulation
%\begin{enumerate}
%\item Given a geometry, compute the shear stress of along the boundary of each of the bodies.
%\item Given the shear stress, update the geometry according to the erosion model.
%\end{enumerate}

@@

For the sake of numerical stability, we will modify the erosion law \eqref{eqn:bodyEvolution} to include a term that depends on the local curvature $\kappa$. The idea is to prevent regions of very high curvature. The modified normal velocity is
in \eqref{eqn:bodyEvolution} as follows
\begin{equation}
\label{eqn:Vn}
\Vn = \atau + \eps \mean{\atau} \left( \frac{L }{2\pi} \kappa - 1 \right)
\end{equation}
Here, $\eps$ is a small parameter, and $L$ is the total arc length of the body, and $\mean{\cdot}$ indicates a surface mean. The second term on the right acts to smooth bodies with high-frequency oscillations by introducing a {\em curvature}-dependent component in the interface velocity. Since a closed body has a mean curvature of $\mean{\kappa} = 2\pi / L$, the term in parentheses has mean zero. Thus, this term is area-preserving, so that the only material loss is due to the shear stress itself. In addition, we have scaled this term on the $\mean{\atau}$, the average absolute stress, so that it is always proportional to the true source of erosion $\atau$, even as bodies vanish (reword).

@@

With the shear stress computed, we are now ready to evolve the interface of each body. As described in Section \ref{sec:thetaL}, we use the {\thL} framework, with $\theta$ being the tangent angle and $L$ the total arc length of a given boundary, and with the boundaries parameterized by normalized arc length $\alpha = 2 \pi s / L$.

In addition to the curvature-dependent term in \eqref{eqn:Vn}, we regularize boundary evolution by applying a narrow Gaussian filter to absolute shear stress
\begin{equation}
\atausig = \int_{\gamma} \frac{1}{\sqrt{2\pi} \sigma} \exp \left(- \frac{\alpha'^2}{2 \sigma^2}\right) \abs{ \tau(\alpha - \alpha') } \, d\alpha'
\end{equation}
Here $\sigma$ is the small parameter that determines the width of the filter. The corresponding normal and tangential interface velocities are the given by
\begin{align}
& \Vnsig = \atausig +  \eps \mean{\atausig}   \left(\pderiv{\theta}{\alpha} - 1 \right) \\
& \pderiv{\Vssig}{\alpha} = \thalpha \Vnsig - \mean{\thalpha \Vnsig}
\end{align}

%The interface evolution is then given by
%\begin{equation}
%\dot{\xx}(t) = \Vn \nn + \Vs \ss.
%\end{equation}

For simplicity, we introduce the variables
\begin{align}
\label{eq:M}
& \Mterm = - 2 \pi \mean{\thalpha \Vnsig} \\
\label{eq:N}
& \NLterm = \frac{2 \pi}{L} \left( \pderiv{\atausig}{\alpha} + \thalpha \Vssig \right) \\
\label{eq:elfun}
& \elfun = \frac{2 \pi  \mean{\atausig}}{L}
\end{align}
So that evolution equations \eqref{eqn:Levo1} and \eqref{eqn:thetaevo1} can be compactly expressed as
\begin{align}
\label{eqn:Levo2}
& \tderiv{L}{t} = \Mterm \\
\label{eqn:thetaevo2}
& \pderiv{\theta}{t} = \eps \elfun \ppd{\theta}{\alpha} + \NLterm
\end{align}

Here $\NLterm$ represents the nonlinear terms (and nonlocal etc) in the PDE for $\theta$.
We will evolve $\theta$ in spectral space, and so introduce the Fourier series
\begin{align}
& \theta(\alpha,t) = \alpha + \FourierSum \thhat_k e^{ i k \alpha} \\
& \NLterm(\alpha,t)  = \FourierSum \hat{\NLterm}_k e^{ i k \alpha}
\end{align}
The careful reader will notice that we have separated out the linear part of $\theta$, so that the remaining function is smooth and has a rapidly decaying Fourier series.
In spectral space, \eqref{eqn:thetaevo2} produces the system of ODEs
\begin{equation}
\tderiv{\thhat_k}{t} +  \eps k^2  \elfun \thhat_k = \hat{\NLterm}_k
\end{equation}
Each ODE is first-order and linear (though variable-coefficient due to $\elfun$) and thus can be solved by using the integrating factor
\begin{equation}
\label{eq:mu}
\mu_k = \exp \left( \eps k^2 \int \elfun \, dt \right)
\end{equation}
We are therefore led to the coupled system of ODEs
\begin{align}
\label{eqn:Levo2}
& \tderiv{L}{t} = \Mterm \\
\label{eqn:thetaevo2}
& \tderiv{}{t}\left( \mu_k \thhat_k \right) = \mu_k \hat{\NLterm}_k
\end{align}
We solve system \eqref{eqn:Levo2}--\eqref{eqn:thetaevo2} with an explicit, second-order Runge-Kutta method (RK2). In particular, we choose the {\em midpoint} RK2 method so as to maximize the smoothing effect of the curvature term.

@@

Next, at the second stage of RK2, we discretize as
\begin{align}
& \frac{L^{(n+1)} - L^{(n)}}{\Dt} = \Mterm^{(n+1/2)} \\
& \frac{ \mu_k^{(n+1)} \thhat_k^{(n+1)} - \thhat_k^{(n)}}{\Dt} = \mu_k^{(n+1/2)} \hat{\NLterm}_k^{(n+1/2)}
\end{align}
We again compute the integrating factor, $\mu_k^{(n+1)}$ via the ( insert integration rule), giving
\begin{align}
\thhat_k^{(n+1)} = 
& \thhat_k^{(n)}
\exp \left( - \epsilon k^2 \Dt \elfun^{(n+1/2)} \right) +  \\
& \Dt \, \hat{\NLterm}_k^{(n+1/2)}
\exp \left( - \frac{1}{4} \eps k^2 \Dt \left( 3 \elfun^{(n+1/2)} - \elfun^{(n)} \right) \right)
\end{align}
